---
title: "EpiSmokEr - Epigenetic Smoking status Estimator- R package"
author: "Sailalitha Bollepalli"
date: "Last updated on: `r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: epismoker.bib
fontsize: 16pt
vignette: >
  %\VignetteIndexEntry{EpiSmokEr - Epigenetic Smoking status Estimator- R package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
***
Several independent studies have reported the high impact of smoking on DNA methylation with current, never and former smokers exhibiting different methylation profiles. Recently, usage of DNA methylation as a proxy measure to estimate smoking status has been proposed [@elliott]. **EpiSmokEr** serves as a comprehensive package with four options to estimate smoking status based on methylation data.

This vignette gives an overview of how to use the **EpiSmokEr** R package. 

## Four options in EpiSmokEr
**EpiSmokEr** provides four options to calculate smoking score and predicts smoking status from whole blood Infinium HumanMethylation450 data.

1. **Multinomial LASSO method (MLM)** provides smoking probabilities for each individual and **predicts smoking status** of the individual. MLM uses weights of 121 CpGs selected by multinomial LASSO approach for each class of smoking status i.e current, former and never smokers. In addition to these 121 CpGs, sex and intercept terms are also used in the calculation.
These coefficients selected by MLM can be viewed as follows:
```{r}
load(system.file("extdata", "MLM_coefficients.rda", package= "EpiSmokEr"))
knitr::kable(head(MLM_coefficients, caption = "121 CpGs and coefficients selected by Multinomial LASSO approach"))
```

2. **Elliott method (EM)** provides smoking score calculated based on the approach outlined in [@elliott]. Elliott *et al* computed smoking scores using 187 smoking-associated CpGs identified in an EWAS comparing current Vs never smokers [@zellinger]. For a detailed explanation of smoking score calculation please refer to [@elliott]. Briefly, weights and reference beta values are obtained from Zellinger *et al*. For each of the 187 CpG sites, the absolute difference between cohort beta value and reference beta value is calculated and then multiplied by the corresponding weight. A smoking score is then obtained by summation across all the 187 CpG site scores.

3. **Zhang method (ZM)** provides methylation score calculated using 4 CpGs.
Zhang *et al * developed the methylation score for discriminating current smokers from never smokers, and former smokers from never smokers using a stepwise selection of logistic regression models [@zhang]. Weights of 4 CpGs provided by Zhang *et al* were used to obtain methylation scores.

4. **all** a comprehensive approach which provides results from all the three methods described above.

## Features of **EpiSmokEr**
- Can perform smoking score calculation starting from idat files.
- Normalization of datasets using channel and colour specific quantile normalization.
- Smoking score calculation using three methods.
- Prediction of smoking status based on the whole blood methylation data.
- Results provided in html and csv format.

# Installation
***
To install EpiSmokEr, start R and then type the following commands:
```{r eval=FALSE}
source("http://bioconductor.org/biocLite.R")
install.packages("devtools") # if you don't have the package, run install.packages("devtools")
library(devtools)
install_github("sailalithabollepalli/EpiSmokEr")
```
**EpiSmokEr** depends on the following packages:
```{r eval=FALSE}
library(IlluminaHumanMethylation450kmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(minfi)
library(htmlTable)
library(rmarkdown)
```
After installing **EpiSmokEr**, load the package along with its dependencies:
```{r}
suppressPackageStartupMessages({
library(EpiSmokEr)  
library(IlluminaHumanMethylation450kmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(minfi)
library(htmlTable)
library(rmarkdown)
})
```

# Input Data
***
To generate the smoking scores from the whole blood methylaion, EpiSmokEr simply requires methylation data as an input.
Input data could be either raw methylation data in the form of idat files or a normalised methylation matrix in the beta scale ranging in between 0 and 1.
For MLM method, a sample sheet with **sex** status is needed to complement the methylation data.

To demonstrate the working of the EpiSmokEr we chose a peripheral blood leukocytes dataset from GEO [GSE42861](http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=gse42861)[@geo]. We use only 6 samples from this dataset to minimise the package size and running time.

#### Following data objects are provided along with the package and are used in the examples below.
- Raw data: idat files of 6 subjects from GSE42861.
- Methylation matrix: A subset of 1000 CpG probes from the normalised methylation matrix (which contained all the CpGs required by EpiSmokEr).
- A sample sheet with gender information

# Smoking Score calculation and predicton of smoking status
*** 
We will now show 2 ways of providing input data to predict smoking status:

## A. From idat files
Here input data are in the form of idat files. `minfi`[@minfi] package is used to read the idat files.
**normaliseData** function has a suite of customised internal functions to normalise and calculate beta values from the idat files.
Color and channel specific normalisation was performed on the training dataset used in the MLM approach. Quantiles from this training dataset are used to adjust the distribution of your input data comparable to the training dataset. Following the normalisation, beta values are calculated.

### Please note:
- Sample sheet must be included in the same folder as the idat files.
- Rownames of samplesheet must match the names of the samples i.e idat files, to facilitate matching of phenotype data with the corresponding methylation data.
- For MLM method, sample sheet should include a column with gender information marked as **sex**, in the format of 1 and 2 representing men and women respectively.

```{r}
rawdata <- loadData(idatPath = system.file("extdata", package = "EpiSmokEr"))
samplesheet <-read.csv(system.file("extdata", "samplesheet_GSE42861.csv", package="EpiSmokEr"), header=TRUE, sep=",")
knitr::kable(head(samplesheet, 5))
```

When raw methylation data is available (idat files) then it is highly recommended to use the same normalisation method implemented in the original article which proposed that specific smoking-score method. All the three smoking score methods (MLM, ZM and EM) we have considered here uses 3 different normalisation methods (QN, SQN and ILM). Therefore we provide three normalisation options to normalise the datasets accordingly and then calculate the smoking score following that specific method.

Quantiles saved from color, channel and probe based quantile normalisation implemented on the training dataset will be used to normalise data for MLM.
```{r}
dataset_QN <- normalizeData(RGset=rawdata, normMethod = "QN") 
```

To use EM, dataset is normalised using subset quantile normalization [@tost]. 
```{r}
dataset_SQN <- normalizeData(RGset=rawdata, normMethod = "SQN")
```

To use ZM, data were normalized to internal controls provided by Illumina (Illumina normalization) and no additional background correction was done.
```{r}
dataset_ILM <- normalizeData(RGset=rawdata, normMethod = "ILM")
```

All the three normalisation methods can be performed using option "ALL". However, the result **dataset_ALL** is saved as a list.
```{r}
dataset_ALL <- normalizeData(RGset=rawdata, normMethod = "ALL")
str(dataset_ALL)
```

Individual normalised datasets can be obtained from **dataset_ALL** as follows:
```{r}
dataset_QN <- dataset_ALL$dataset_QN
dataset_ILM <- dataset_ALL$dataset_ILM
dataset_SQN <- dataset_ALL$dataset_SQN
```

Now the **dataset** object can be passed on to the **epismoker** function to calculate smoking score and smoking status using four options available.


Once we have normalised the methylation data, we can then proceed with smoking status estimation. 
EpiSmokEr provides four options:

### 1. Multinomial LASSO method (MLM)
In this method weights of 121 CpGs selected by Multinomial LASSO approach, and sex and intercept terms are used in the calculation of smoking probabilities for each individual. We get smoking probabilities for the three smoking status classes, i.e current, former and never smokers. Predicted smoking status of an individual is the class having the highest probability. 
```{r}
result_MLM <- epismoker(dataset=dataset_QN, samplesheet = samplesheet, method = "MLM")
result_MLM
# PredictedSmokingStatus represents the smoking status class assigned to the individual based on the calculated smoking probabilities.
```

### 2. Elliott method (EM)
In this method 187 CpGs are used to calculate smoking score as expained in [@elliott].
```{r}
result_EM <- epismoker(dataset = dataset_SQN, method = "EM")
result_EM
```

### 3. Zhang method (ZM)
In this method 4 CpGs are used to calculate methylation score as expained in [@zhang].
```{r}
result_ZM <- epismoker(dataset = dataset_ILM,  method = "ZM")
result_ZM
```

### 4. Compreshensive approach (all)
All three methods are run in this option and results are integrated into a single result object.
```{r}
result_All <- epismoker(dataset_QN=dataset_QN, dataset_ILM=dataset_ILM, dataset_SQN=dataset_SQN, samplesheet = samplesheet, method = "all")
result_All
```

## B. From methylation matrix
A data matrix of normalised methylation values in beta scale, with rows labelling the CpG probe ids and columns labelling the sample names, can also be used as an input to EpiSmokEr. Here we use the methylation matrix from the same 6 samples to demonstrate how methylation matrix can be used as input. 
```{r, include= FALSE}
data("dummyBetaData")
```
Methylation matrix looks likes this. Methylation values are in beta scale ranging betweeen 0 and 1.
```{r}
head(dummyBetaData)
```
Samplesheet must contain the column **sex**, in the format of 1 and 2 representing men and women respectively.
```{r, echo=FALSE, results='asis'}
samplesheet <- read.csv(system.file("extdata", "samplesheet_GSE42861.csv", package="EpiSmokEr"), header=TRUE, sep=",")
knitr::kable(head(samplesheet, 5))
```
```{r}
result <- epismoker(dataset=dummyBetaData, samplesheet = samplesheet, method = "MLM")
knitr::kable(head(result, 5))
```

# Generate Results
***
Results can be saved as html and csv files using **generateReport** function. A **Results** folder will be created in the **current working directory**. If **Results** folder already exists, result files will be appended to this folder.

**generateReport** takes **result object (resObj)** and **method** used in epismoker function. Users have to provide a name for the output html file. 

#### Please note that
* In the R environment, result object from **epismoker** function must be saved using variable names **result_MLM**, **result_EM**, **result_ZM** and **result_All** based on the option used. This enables generation of html report.
   
* In order to create the **Results** folder, set the working directory path to the folder where you have permission to write.

In the example below result object from the **all** option of **epismoker** function has been saved to **result** variable, which will be passed onto generateReport function. On executing **generateReport** function, a **Results** folder will be created in the current working directory, with a html file named "Results_comprehensive.html". Result files generated from the example below can be found in the **Results** folder of the package.

```{r}
setwd(system.file(package = "EpiSmokEr")) # set the directory where you want the result files to be saved
generateReport(resObj = result_All, outputFileName = "Results_comprehensive", method = "all")
```

# References 
***
